Dynamic Prediction of Non-Guassian Outcome with fast Generalized Functional Principal Analysis

### 2023-04-01 Write-ups/Manuscript.Rmd
discussion

<!-- Precision of prediction, usually quantified by the variance of prediction error $Var(\hat{\eta_i}(s)-\eta_i(s))$, is also straightforward under this framework. In step 4 we calculated the MLE of $\hat{\xi}_{ik}$. Based on likelihood theory, its variance can be estimated with observed information ${I}(\hat{\xi}_{ik})$, which is essentially the second derivative of likelihood at  $\hat{\xi}_{ik}$. Therefore, the variation of prediction interval is: -->

<!-- $$Var(\hat{\eta_i}(s)-\eta_i(s))=\boldsymbol{\Phi}(s)I(\hat{\boldsymbol{\xi}}_i)\boldsymbol{\Phi}^T(s)+\hat{\sigma_{\epsilon}}^2$$ -->
<!-- Where $\boldsymbol{\Phi}(s)=(\phi_1(s)...\phi_K(s))$ and $\hat{\boldsymbol{\xi}}_i=(\hat{\xi}_{i1}, ...,\hat{\xi}_{iK})$. -->

@@

```{r AUC_fGFPCA, cache=TRUE}
# maximum observations time
t_vec <- c(200, 400, 600, 800)

@@

We will illustrate the performance of our proposed method using data from a randomized trial on weight loss at University of Colorado Anschutz Medical Campus [@]. The study originally aimed to study the effect of dietary strategies on weight loss, and participants were given a smart scale for daily at-home weighing. This record was then scaled into a binary indicator of participant adherence to the daily-weigh-in procedure, where 1 indicates participants followed the requirement and 0 otherwise. In this section, we will use this series of daily indicators as the binary functional outcome, and use the proposed method to dynamically predict participant adherence on future days given historical track.

@@

```{r AUC_fGFPCA_Appl, cache=TRUE}
# maximum observations time
t_vec2 <- c(100, 200, 300)

@@

```
Discussion
=================================================================================================================================
-	Grid 
-	Score bias: cannot demonstrate without repeat simulation

>>>>

@@

```{r AUC_fGFPCA}
# maximum observations time
t_vec <- c(200, 400, 600, 800)

@@

We will illustrate the performance of our proposed method using data from a randomized trial on weight loss at University of Colorado Anschutz Medical Campus [@bothwell2022; @ostendorf2022]. The study originally aimed to study the effect of dietary strategies on weight loss, and participants were given a smart scale for daily at-home weighing. This record was then scaled into a binary indicator of participant adherence to the daily-weigh-in procedure, where 1 indicates participants followed the requirement and 0 otherwise. In this section, we will use this series of daily indicators as the binary functional outcome, and use the proposed method to dynamically predict participant adherence on future days given historical track.

@@

```{r AUC_fGFPCA_Appl}
# maximum observations time
t_vec2 <- c(100, 200, 300)

@@

```
Discussion
=================================================================================================================================

The simulation study and case study of daily weigh-in data above have revealed the proposed method can achieve better predictive performance with much less computational burden. Compared to existing method based on linear mixed model, fGFPCA can accomodate flexible correlation struture between repeated measure, as well as high density of measurement. This is particularly useful when underlying outcome shows cyclic pattern, such as the simulation setting in Section 3. While the binary outcomes are generated from a cyclic Gaussian process, fGFPCA has much lower ISE and higher AUC than GLMM. Its prediction accuracy also improves as more outcomes are observed, which is a desirable property for dynamic prediction where the datasets would be periodically updated with newly collected data.

While the proposed method achieved good accuracy of point prediction, it is more complicated to develop a straightforward, interpretable metric to evaluate the precision of prediction. The challenge arises from the fact that fitting local GLMMs (Step 2) and FPCA (Step 3) can both introduce uncertainty to the final prediction. Since they are implemented consecutively, it is not clear how the uncertainty from both procedures should be integrated into one single interval estimator. In practice, one could consider developing interval predictions conditional on the local GLMMs, quantifying uncertainty of individual score estimates from FPCA model alone. Since the score $\hat{\xi}_{ik}$ is essentially a maximum likelihood estimate, it is natural to estimate its variance based on likelihood theory with observed information ${I}(\hat{\xi}_{ik})$. Therefore, the variation of prediction interval is:

$$Var(\hat{\eta_i}(s)-\eta_i(s))=\boldsymbol{\Phi}(s)I(\hat{\boldsymbol{\xi}}_i)\boldsymbol{\Phi}^T(s)+\hat{\sigma_{\epsilon}}^2$$
Where $\boldsymbol{\Phi}(s)=(\phi_1(s)...\phi_K(s))$ and $\hat{\boldsymbol{\xi}}_i=(\hat{\xi}_{i1}, ...,\hat{\xi}_{iK})$.

The hyperparameters introduced by the binning procedure in Step 1 can also affect the final predictive performance, thus further investigations are needed to quanitfy such effects. Predictive accuracy and computational speed can change with bin width, number of observations in each bin, whether the bins overlap with each other, also whether the bins are equally spaced along the functional grid. Moreover, if the bins are very narrow with only a few observations, unidentifiability issues can happen when fitting local mixed models. While this may not necessarily affect point predictions, it can cause numeric failure to estimate standard error and prediction intervals. In practice, we expect this to happen when the measurements are sparse or sample size is small. Under such scenarios, one may choose traditional mixed model framework instead to get valid estimates at the expense of computational time.

It should also be noted that the binning procedure in Step 1 changes the density of repeated measures. The final prediction of latent continuous function tracks are made on the binned grid, integrating information contained in several observations in the same bin. In Section 3 and 4, we have transformed this sparser grid to the original, unbinned grid using linear interpolation of the predicted latent function. However, this is by no means the only method to be considered. Such transformation can also be made by projection to a higher-dimensional space, similar as general additive model with a large number of basis functions. We suspect that this method would work better than linear interpolation when the original grid is sparse or irregular, but take longer time for computation.

<!-- -	Score bias: cannot demonstrate without repeat simulation -->

### 2023-04-01 Write-ups/Manuscript.tex
discussion

\title{Dynamic Prediction of Non-Guassian Outcome with fast Generalized
Functional Principal Analysis}
\author{Ying Jin \and Andrew Leroux}
\date{March 31, 2023}

@@

We will illustrate the performance of our proposed method using data
from a randomized trial on weight loss at University of Colorado
Anschutz Medical Campus {[}@{]}. The study originally aimed to study the
effect of dietary strategies on weight loss, and participants were given
a smart scale for daily at-home weighing. This record was then scaled
into a binary indicator of participant adherence to the daily-weigh-in
procedure, where 1 indicates participants followed the requirement and 0
otherwise. In this section, we will use this series of daily indicators
as the binary functional outcome, and use the proposed method to
dynamically predict participant adherence on future days given
historical track.

@@

\begin{itemize}
\tightlist
\item
  Grid
\item
  Score bias: cannot demonstrate without repeat simulation
\end{itemize}

@@

\hypertarget{refs}{}
\begin{CSLReferences}{1}{0}
\leavevmode\vadjust pre{\hypertarget{ref-chen2013}{}}%
Chen, H., Wang, Y., Paik, M. cho, and Choi, H. A. (2013), {``A marginal
approach to reduced-rank penalized spline smoothing with application to
multilevel functional data,''} \emph{J Am Stat Assoc.}, 108, 1216--1229.
\url{https://doi.org/10.1080/01621459.2013.826134}.

@@

>>>>

\title{Dynamic Prediction of Non-Guassian Outcome with fast Generalized
Functional Principal Analysis}
\author{Ying Jin \and Andrew Leroux}
\date{April 01, 2023}

@@

We will illustrate the performance of our proposed method using data
from a randomized trial on weight loss at University of Colorado
Anschutz Medical Campus (\protect\hyperlink{ref-bothwell2022}{Bothwell
et al. 2022}; \protect\hyperlink{ref-ostendorf2022}{Ostendorf et al.
2022}). The study originally aimed to study the effect of dietary
strategies on weight loss, and participants were given a smart scale for
daily at-home weighing. This record was then scaled into a binary
indicator of participant adherence to the daily-weigh-in procedure,
where 1 indicates participants followed the requirement and 0 otherwise.
In this section, we will use this series of daily indicators as the
binary functional outcome, and use the proposed method to dynamically
predict participant adherence on future days given historical track.

@@

The simulation study and case study of daily weigh-in data above have
revealed the proposed method can achieve better predictive performance
with much less computational burden. Compared to existing method based
on linear mixed model, fGFPCA can accomodate flexible correlation
struture between repeated measure, as well as high density of
measurement. This is particularly useful when underlying outcome shows
cyclic pattern, such as the simulation setting in Section 3. While the
binary outcomes are generated from a cyclic Gaussian process, fGFPCA has
much lower ISE and higher AUC than GLMM. Its prediction accuracy also
improves as more outcomes are observed, which is a desirable property
for dynamic prediction where the datasets would be periodically updated
with newly collected data.

While the proposed method achieved good accuracy of point prediction, it
is more complicated to develop a straightforward, interpretable metric
to evaluate the precision of prediction. The challenge arises from the
fact that fitting local GLMMs (Step 2) and FPCA (Step 3) can both
introduce uncertainty to the final prediction. Since they are
implemented consecutively, it is not clear how the uncertainty from both
procedures should be integrated into one single interval estimator. In
practice, one could consider developing interval predictions conditional
on the local GLMMs, quantifying uncertainty of individual score
estimates from FPCA model alone. Since the score \(\hat{\xi}_{ik}\) is
essentially a maximum likelihood estimate, it is natural to estimate its
variance based on likelihood theory with observed information
\({I}(\hat{\xi}_{ik})\). Therefore, the variation of prediction interval
is:

\[Var(\hat{\eta_i}(s)-\eta_i(s))=\boldsymbol{\Phi}(s)I(\hat{\boldsymbol{\xi}}_i)\boldsymbol{\Phi}^T(s)+\hat{\sigma_{\epsilon}}^2\]
Where \(\boldsymbol{\Phi}(s)=(\phi_1(s)...\phi_K(s))\) and
\(\hat{\boldsymbol{\xi}}_i=(\hat{\xi}_{i1}, ...,\hat{\xi}_{iK})\).

The hyperparameters introduced by the binning procedure in Step 1 can
also affect the final predictive performance, thus further
investigations are needed to quanitfy such effects. Predictive accuracy
and computational speed can change with bin width, number of
observations in each bin, whether the bins overlap with each other, also
whether the bins are equally spaced along the functional grid. Moreover,
if the bins are very narrow with only a few observations,
unidentifiability issues can happen when fitting local mixed models.
While this may not necessarily affect point predictions, it can cause
numeric failure to estimate standard error and prediction intervals. In
practice, we expect this to happen when the measurements are sparse or
sample size is small. Under such scenarios, one may choose traditional
mixed model framework instead to get valid estimates at the expense of
computational time.

It should also be noted that the binning procedure in Step 1 changes the
density of repeated measures. The final prediction of latent continuous
function tracks are made on the binned grid, integrating information
contained in several observations in the same bin. In Section 3 and 4,
we have transformed this sparser grid to the original, unbinned grid
using linear interpolation of the predicted latent function. However,
this is by no means the only method to be considered. Such
transformation can also be made by projection to a higher-dimensional
space, similar as general additive model with a large number of basis
functions. We suspect that this method would work better than linear
interpolation when the original grid is sparse or irregular, but take
longer time for computation.

@@

\hypertarget{refs}{}
\begin{CSLReferences}{1}{0}
\leavevmode\vadjust pre{\hypertarget{ref-bothwell2022}{}}%
Bothwell, S., Kaizer, A., Catenacci, V., and Wrobel, J. (2022),
{``Pattern-based clustering of daily weigh-in trajectories using dynamic
time warping,''} \emph{Biometrics}.
\url{https://doi.org/10.1111/biom.13773}.

\leavevmode\vadjust pre{\hypertarget{ref-chen2013}{}}%
Chen, H., Wang, Y., Paik, M. cho, and Choi, H. A. (2013), {``A marginal
approach to reduced-rank penalized spline smoothing with application to
multilevel functional data,''} \emph{J Am Stat Assoc.}, 108, 1216--1229.
\url{https://doi.org/10.1080/01621459.2013.826134}.

@@

\leavevmode\vadjust pre{\hypertarget{ref-ostendorf2022}{}}%
Ostendorf, D. M., Caldwell, A. E., Zaman, A., Pan, Z., Bing, K.,
Wayland, L. T., Creasy, S. A., Bessesen, D. H., MacLean, P., Melanson,
E. L., and Catenacci, V. A. (2022), {``Comparison of weight loss induced
by daily caloric restriction versus intermittent fasting (DRIFT) in
individuals with obesity: Study protocol for a 52-week randomized
clinical trial,''} \emph{Trials}, 23, 718.
\url{https://doi.org/10.1186/s13063-022-06523-2.}
